R Markdown file for Reproducible Research Week 2
=================================================

##First steps is to load the data and look at some basic summary statistics of the data 
```{r review criteria 1, echo = TRUE}
setwd("//LanFspd01.ncsbn.org/UserProfileFolders$/LErickson/Documents/Coursera/data")
Activity <- read.csv("activity.csv", header = TRUE, na.strings="NA")
summary(Activity)
rdate <- as.Date(Activity$date,"%m/%d/%y")
head(Activity)
```



##Now, I want to look at a **histogram** of the total number of steps taken each day. 
To do this, I need to:

          1. Calculate the sum of the steps taken each day
          2. Create the histogram of the calculated steps

```{r review criteria 2, echo = TRUE}
TotalDailySteps <- tapply(Activity$steps, Activity$date, FUN = sum, na.rm = TRUE) 
hist(TotalDailySteps, breaks=10, xlab="Steps per Day", col="pink", main="Total Number of Steps Taken Each Day")
```



##Next, I want to to know what the **mean** and **median** number of steps taken each day and rounded to one digit
```{r review criteria 3, echo = TRUE}
options(digits=1)  ##round to one digit
stepsmean <- mean(TotalDailySteps, na.rm = TRUE)
stepsmedian <- median(TotalDailySteps, na.rm = TRUE)
```
The mean number of steps taken each day is `r stepsmean`.  
The median number of steps taken each day is `r stepsmedian`.



##I want to view a time series plot of the average number of steps taken per interval.  
To do this, I need to:

                   1. Calculate the mean number of steps per interval
                   2. Create a data frame with the caluclated mean
                   3. Create the plot of the new data frame as a time series plot
                   
```{r review criteria 4, echo = TRUE}
IntervalStep <- tapply(Activity$steps, Activity$interval, mean, na.rm = TRUE) ##Calculating the mean number of steps taken at each interval
IntAvg <- data.frame(interval=as.integer(names(IntervalStep)), avg=IntervalStep) ##creating a data frame from IntervalStep
plot(IntAvg$avg ~ IntAvg$interval, type='l', col="pink", xlab = "Interval", ylab = "Average Steps", main="Average Steps per Interval")
```



##Next, I want to find the maximum number of steps taken during a 5-minute interval
```{r review criteria 5, echo = TRUE}
StepsMax <- IntAvg[which.max(IntAvg$avg), ]
```
The maximum number of steps taken during a 5-minute interval is `r StepsMax$avg` steps at interval `r StepsMax$interval`



##This code shows how I dealt with missing data
first step is to calculate the number of missing data values
```{r review criteria 6.1, echo = TRUE}
MissingSteps <- sum(is.na(Activity$steps))  ##to find how many missing data points there are
```

There are `r MissingSteps` missing data values


Next, I need to apply the median for a 5-minute interval and use that number of fill the missing data values. Here are my steps to do this:

        1. Calculate the median number of steps for each interval. 
        2. Create a data frame from the median calculated. 
        3. Merge the created data frame with the original data set. 
        4. Replace the missing values with the calculated median. 

```{r review criteria 6.2, echo = TRUE}
MedStep <- tapply(Activity$steps, Activity$interval, median, na.rm = TRUE) ##Calculating the mean number of steps taken at each interval
IntMed <- data.frame(interval=as.integer(names(MedStep)), step=MedStep) ##creating a data frame from MedStep

TempActivity <- merge(Activity, IntMed, by="interval", all.y = FALSE) ##Merge the created data frame with the original data set
TempActivity$steps[is.na(TempActivity$steps)] <- as.integer(round(TempActivity$step[is.na(TempActivity$steps)])) ##uses the calculated median in place of the missing values
keeps <- names(Activity)
TempActivity <- TempActivity[keeps]
MissingSteps2 <- sum(is.na(TempActivity$steps)) 
```
We can see that this method has producted `r MissingSteps2` missing data values


##Let's look at a histogram of the total steps taken where the missing data is replaced with the median of the 5-minute interval
To do this, I need to:

          1. Calculate the total steps per each day
          2. create a histogram of this calculation

```{r review criteria 7, echo = TRUE}
RevTotalDailySteps <- tapply(TempActivity$steps, TempActivity$date, FUN = sum, na.rm = TRUE) 
hist(RevTotalDailySteps, breaks=10, main="Total Steps using Median for NAs", xlab="Steps per Day", col="pink")
```


##I want to look at the differences in average number of steps taken per 5-minute interval across weekdays and weekends

      1. Create a variable that labels each oberservation as a weekend or weekday based on the given date
      2. Calculate the mean number of steps per each interval
      3. Load lattice because we want to stack two graphs
      4. Plote the data


```{r review criteria 8, echo = TRUE}
Weekend <- c("Saturday", "Sunday") #Identifying the weekend
TempActivity$weekday = as.factor(ifelse(is.element(weekdays(as.Date(TempActivity$date)), Weekend), "Weekend",  "Weekday"))  #flesh out the weekends by using an if else statement - if it is not a weekend, it is a weekday


IntervalDow <- aggregate(steps ~ interval + weekday, TempActivity, mean)  
head(IntervalDow)

library(lattice)
xyplot(IntervalDow$steps ~ IntervalDow$interval|IntervalDow$weekday, layout=c(1,2), type="l", col="pink", main="Steps per Interval Comparing Weekday to Weekend", ylab="Steps", xlab="Interval")
```
